---
title: "Practice with dplyr and ggplot2"
author: "Dr. Shoemaker"
date: "DATA 2401"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Fancy Histograms!

We'll start by making some histograms.

```{r}
#install.packages("dslabs")
library(dslabs) 
data(heights)
glimpse(heights)
```

This data is the heights of humans, divided by their biological sex.

Use ggplot to make a histogram of all of the heights:

```{r}
ggplot(heights, aes(x = height)) + geom_histogram()

```

Change up the binwidth and see how the plots change. Try 1, 5, 10, and 20

```{r}
ggplot(heights, aes(x = height)) + geom_histogram(binwidth = 1)
ggplot(heights, aes(x = height)) + geom_histogram(binwidth = 5)
ggplot(heights, aes(x = height)) + geom_histogram(binwidth = 10)
ggplot(heights, aes(x = height)) + geom_histogram(binwidth = 20)

```

Smooth this out to an emperical density with `geom_density()`

```{r}
ggplot(heights, aes(x = height)) + geom_density()
```

Use a new argument in the `aes()`, `group =` to split this density by sex

```{r}
ggplot(heights, aes(x = height, group = sex)) + geom_density()
```

OR we can do it with `color` or `fill.` If you say you want to color by sex, R knows that you want a different curve for each of them.

```{r}
ggplot(heights, aes(x = height, color = sex)) + geom_density()


```

If you've used fill, then there is now a slight issue that they are overlapped. We can fix this with alpha transparency!

```{r}
ggplot(heights, aes(x = height, fill = sex)) + geom_density(alpha = 0.5)


```

Let's make some boxplots of the same information.

```{r}
ggplot(heights, aes(x = sex, y = height)) + geom_boxplot()
```

## Quantatitive summaries:

Find the mean and median overall.

```{r}
ggplot(heights, aes(x = sex, y = height)) + geom_boxplot()
heights %>% summarize(mean_height = mean(height), median_hieght = median(height))

```

Find the mean and median for both groups.

```{r}
heights %>% group_by(sex) %>%
  summarize( mean_height = mean(height), median_height = median(height))
```

How tall is the tallest woman? How short is the shortest man?

```{R}
heights %>% filter(sex == "Female") %>% summarise(tallest = max(height))
heights %>% filter(sex == "Male") %>% summarise(shortest = min(height))


```

Use your summary table and pipe it to a ggplot to make a bar graph showing the mean of both groups.

```{R}
heights %>%
  group_by(sex) %>% summarize(mean_height = mean(height)) %>% ggplot(aes(x = sex, y = mean_height, fill = sex))+ geom_col()
```

## Presidental Elections Data

```{r}
# install.packages("pscl")
library(pscl) # loads in the package that has this data. 
## You might need to install this...

# data for presidental elections
votedata <-  presidentialElections
glimpse(votedata)
```

Let's look at the democratic vote by state for 2000. We can't use `geom_bar` for a bar chart, since we have the category in one variable and the "height" of the bar in another. We need `geom_col()`

Make a bar graph of the democratic vote by state in 2000.

```{r}
votedata %>% filter(year == 2000) %>% ggplot(aes(x = state, y = demVote)) +
  geom_col()

```

Well this looks awful. We have two options: swap the x and y or the more fun sounding... Coordinate flip!

Use `coord_flip()` on the previous graph to make it better.

```{r fig.height=7, fig.width=5}
votedata %>% filter(year == 2000) %>% ggplot(aes(x = state, y = demVote)) +
geom_col()
coord_flip()

```

I don't love the squashed together coordinates, but it's a display window issue.

So. This is a helpful graph, but it would be *more* helpful if it was ordered. Use `x = reorder(x_variable, y_variable)` in `aes()` to order the x variable by the y variable

```{r}
votedata %>% filter(year == 2000) %>% ggplot(aes(x = reorder(state, demVote), y = demVote)) +
  geom_col() +
  coord_flip()


```

So, what if I want to see what the north and south states did different?

start with a facet_wrap using the `south` variable:

```{r}
votedata %>%
  filter(year == 2000) %>%
  ggplot(aes(x = reorder(state, demVote), y = demVote)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~south)

```

Okay, that's not great. Lets *color* by south instead.

```{r}
votedata %>%
  filter(year == 2000) %>%
  ggplot(aes(x = reorder(state, demVote), y = demVote, fill = south)) +
  geom_col() +
  coord_flip()

```

I'm a good data scientist, so I want my plot to have a name! and my axes to have lables! Use `labs` to add a title, subtitle, and x and y labels.

```{r, eval = F}
votedata %>% filter(year == 2000) %>%
  ggplot(aes(x = reorder(state, demVote), y = demVote, fill = south)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Democratic Vote by State in 2000",
    subtitle = "Ordered by vote percentage",
    x = "State",
    y = "Democratic Vote %"
  )

```

You can move the legend with `theme(legend.position = "bottom")`

```{r, eval = F}
votedata %>% filter(year == 2000) %>% ggplot(aes(x = reorder(state, demVote), y = demVote, fill = south)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Democratic Vote by State in 2000",
    x = "State",
    y = "Democratic Vote %"
  ) +
  theme(legend.position = "bottom")


```

What else could we facet by? years! Let's filter to year in 2008 and 2016, then facet by years.

```{r}
votedata %>% filter(year %in% c(2008,2016)) %>% ggplot(aes(x = reorder(state, demVote), y = demVote, fill = south)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~year)


```

We need to know who won! We could add a vertical line at 50 for who got more, to indicate the majority of votes. Adding the layer `geom_hline()` adds a horizontal line. (What do you guess `geom_vline()` would do?)

```{r}
votedata %>% filter(year %in% c(2008,2016)) %>% ggplot(aes(x = reorder(state, demVote), y = demVote, fill = south)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~year) +
  geom_hline(yintercept = 50)

```

### Getting fancy with a map!

When using geom_polygon or geom_map, you will typically need two data frames:

-   one contains the coordinates of each polygon (positions)
-   the other the values associated with each polygon (values).

An id variable links the two together.

Run the below example code to get map graphs for votes in 2008 and then 2016.

```{r}
library(maps)
votedata$state <- tolower(votedata$state)  ## states need to be lowercase for linking

states_map <- map_data("state") ## this gives us the lat and long for each point of each state.
  
map_plot <-  ggplot(data =  votedata %>% filter(year == 2008), aes(map_id = state)) +
    geom_map(aes(fill = demVote), map = states_map) +
    expand_limits(x = states_map$long, y = states_map$lat)

map_plot  
```

```{r}
map_plot <-  ggplot(data =  votedata %>% filter(year == 2016), aes(map_id = state)) +
  geom_map(aes(fill = demVote), map = states_map)+
  expand_limits(x = states_map$long, y = states_map$lat)

map_plot  
```

What if I want a map that shows which of the states are "south"? What do I change?

```{r}
map_plot <-  ggplot(data =  votedata %>% filter(year == 2016), aes(map_id = state)) +
  geom_map(aes(fill = south), map = states_map)+
  expand_limits(x = states_map$long, y = states_map$lat)

map_plot

```

### Some more dplyr practice

I want to know the average democratic vote for N vs S, by year.

First, find the average democratic votes for the north and the south, every year. You'll need to do a *double* `group_by()` here. You do it in one call of the function.

```{r}
votedata %>%
  group_by(year, south) %>% summarize(avg_vote = mean(demVote))

```

Then, let's plot that! Pipe the result of your group_by and summarize to ggplot and geom_line(), with year on the x axis and your summarized value on the y axis. Color by the `south` variable.

```{r}
votedata %>% group_by(year, south) %>% summarize(avg_vote = mean(demVote)) %>%
  ggplot(aes(x = year, y = avg_vote, color = south)) +
  geom_line()

```

## Layering plots!

Penguins!

```{r}
library(palmerpenguins)
glimpse(penguins)
```

We can use boxplots to visualize the distribution of weight (body_mass_g) within each species:

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()

```

What if we also want the points? Layering!! Add a geom_point to your existing boxplot. geom_boxplot + geom_point!

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  geom_point()

```

But, these are all stacked up... to actually see them, use "geom_jitter" instead of points

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  geom_jitter()

```

How to get the boxplots on top? The layers are plotted in the order you give them, so change the ordering to geom_point + geom_boxplot. (You might want to change the alpha on the boxplot to be able to see the plots under them)

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_jitter() +
  geom_boxplot(alpha = 0.5)


```

Maybe let's try replacing the boxplot with a `geom_violin()`?

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_jitter() +
  geom_violin(alpha = 0.5)


```

## More Practice with Penguins

-   Count the number of male and female penguins for each species.
-   Find the average penguin mass for each island
-   Find the average bill length by sex
-   Make a plot of bill_length and bill_depth, but only for the female penguins.
-   Make a plot of flipper length and body mass, faceted by island and colored by species.
-   Plot the density of the flipper length, colored by sex
-   Plot the density of the body mass, colored by year

\<or, make your own similar graphs with any of the data we've used in class so far!\>

```{r}
penguins %>% group_by(species, sex) %>%
  count()

penguins %>% group_by(island) %>%
  summarize(avg_mass = mean(body_mass_g, na.rm = T))

penguins %>% group_by(sex) %>%
  summarize(avg_bill_length = mean(bill_length_mm, na.rm = T))

penguins %>% filter(sex == "female") %>%
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point()

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  facet_wrap(~island)

ggplot(penguins, aes(x = flipper_length_mm, fill = sex)) +
  geom_density(alpha = 0.5)

ggplot(penguins, aes(x = body_mass_g, fill = as.factor(year))) +
  geom_density(alpha = 0.5)


```
